---
title: "Class 09 Halloween Candy Mini Proj"
author: "Dylan Mullaney (A16869792)"
format: html
---

#Today we will examine data from 538 on common Halloween candy. In this particular project we will use ggplot, dplyr, and PCA to make sense of the multivariable dataset 

##Importing candy data 

#For a txt file, could use read.table

```{r}
url <- "https://raw.githubusercontent.com/fivethirtyeight/data/master/candy-power-ranking/candy-data.csv"
candy <- read.csv (url, row.names=1)
head(candy)
```

> Q1. How many different candy types are in this dataset?

```{r}
nrow(candy)
```

> Q2. Q2. How many fruity candy types are in the dataset?

```{r}
sum(candy$fruity)
#This is a count value, not a percent
```

#Winpercent refers to the likelihood a candy will be chosen as the winner.

> Q3. What is your favorite candy in the dataset and what is it’s winpercent value?

```{r}
candy["Haribo Happy Cola",] $winpercent
#This is a percent value, not a count
```

> Q4. What is the winpercent value for “Kit Kat”?

```{r}
candy["Kit Kat",] $winpercent
```

> Q5. What is the winpercent value for “Tootsie Roll Snack Bars”?

```{r}
candy["Tootsie Roll Snack Bars",] $winpercent
```

#How many candies are chocolate?

```{r}
sum(candy$chocolate)
```

> Q6. Is there any variable/column that looks to be on a different scale to the majority of the other columns in the dataset?

```{r}
#Install package in R brain
library("skimr")
skim(candy)
```


>Q7. What do you think a zero and one represent for the candy$chocolate column?


> Q8. Plot a histogram of winpercent values


> Q9. Is the distribution of winpercent values symmetrical?


> Q10. Is the center of the distribution above or below 50%?


> Q11. On average is chocolate candy higher or lower ranked than fruit candy?


> Q12. Is this difference statistically significant?

















