--- 
title: "Class 18: Pertussis Mini Project"
author: "Dylan Mullaney (A168696792)"
format: pdf 
---

Pertussis (aka Whooping Cough) is a deadly lung infection caused by the bacteria B. Pertussis.

The CDC tracks Pertussis cases around the US.
tinyurl.com / pertussiscdc

There is interesting data to look at here, but it's not in an excel or csv file to download. We need to scrape the data using the **datapasta** pacakge. https://github.com/MilesMcBain/datapasta

> Q1. With the help of the R “addin” package datapasta assign the CDC pertussis case number data to a data frame called cdc and use ggplot to make a plot of cases numbers over time.

```{r, include="FALSE"}
cdc <-
  data.frame(
    year = c(1922L,1923L,1924L,1925L,1926L,1927L,
             1928L,1929L,1930L,1931L,1932L,1933L,1934L,
             1935L,1936L,1937L,1938L,1939L,1940L,
             1941L,1942L,1943L,1944L,1945L,1946L,
             1947L,1948L,1949L,1950L,1951L,1952L,
             1953L,1954L,1955L,1956L,1957L,1958L,
             1959L,1960L,1961L,1962L,1963L,1964L,
             1965L,1966L,1967L,1968L,1969L,1970L,
             1971L,1972L,1973L,1974L,1975L,1976L,1977L,
             1978L,1979L,1980L,1981L,1982L,1983L,
             1984L,1985L,1986L,1987L,1988L,1989L,
             1990L,1991L,1992L,1993L,1994L,1995L,
             1996L,1997L,1998L,1999L,2000L,2001L,
             2002L,2003L,2004L,2005L,2006L,2007L,
             2008L,2009L,2010L,2011L,2012L,2013L,
             2014L,2015L,2016L,2017L,2018L,2019L,2020L,
             2021L,2022L, 2024L),
    cases = c(107473,164191,165418,152003,202210,181411,
                                     161799,197371,166914,172559,215343,
                                     179135,265269,180518,147237,214652,
                                     227319,103188,183866,222202,191383,191890,
                                     109873,133792,109860,156517,74715,
                                     69479,120718,68687,45030,37129,60886,
                                     62786,31732,28295,32148,40005,14809,
                                     11468,17749,17135,13005,6799,7717,9718,
                                     4810,3285,4249,3036,3287,1759,2402,
                                     1738,1010,2177,2063,1623,1730,1248,
                                     1895,2463,2276,3589,4195,2823,3450,4157,
                                     4570,2719,4083,6586,4617,5137,7796,
                                     6564,7405,7298,7867,7580,9771,11647,
                                     25827,25616,15632,10454,13278,16858,
                                     27550,18719,48277,28639,32971,20762,
                                     17972,18975,15609,18617,6124,2116,3044, 35493)
  )
```

```{r}
head(cdc)
```


Let's plot the new data
```{r}
library(ggplot2)

ggplot(cdc, aes(year, cases))+
  geom_point() +
  geom_line() +
  ylab("Number of Cases")+
  xlab("Year")+
  ggtitle("Cases of Pertussis per Year") 
```


> Q2. Using the ggplot geom_vline() function add lines to your previous plot for the 1946 introduction of the wP vaccine and the 1996 switch to aP vaccine (see example in the hint below). What do you notice?

```{r}
ggplot(cdc, aes(year, cases))+
  geom_point() +
  geom_line() +
  geom_vline(xintercept=1946,col="blue")+
  geom_vline(xintercept=1996,col="red")+
   geom_vline(xintercept=2004,col="gray")+
  geom_vline(xintercept=2020,col="green")+
  ylab("Number of Cases")+
  xlab("Year")+
  ggtitle("Cases of Pertussis per Year") 
```

> Q3. Describe what happened after the introduction of the aP vaccine? Do you have a possible explanation for the observed trend?

There were high case numbers before the first whole cell vaccine (blue line), then a rapid decline in cases following.There was a general hold of low values after the aP vaccine (red line), followed by a spike in 2004s. This could be a point where people began to mistrust vaccines and not get them as frequently or the vaccine could have been reformatted. There is a spike again following the green line after COVID19. 

Infants who got the new vaccine are growing up and their immunity is waning and they need a new booster shot. It seems that something has to be different about the immune response to infection with the older wP vaccine versus the newer aP vaccine. 


## Computational Models of Immunity, Pertussis Booster Shot (CMI-PB)

The CMI-PB project aims to address this key question - what's different between aP and wP vaccinated individuals? This is called systems vaccinology. https://www.cmi-pb.org/

We can get all the data from this ongoing project via JSON API.For this we will use the **jsonlite** package.

```{r}
library(jsonlite)

subject <- read_json ("https://www.cmi-pb.org/api/v5_1/subject", simplifyVector = TRUE) 
head(subject)
```

> Q. How many individual subjects are in this dataset?

```{r}
nrow(subject)
```

> Q4. How many aP and wP infancy vaccinated subjects are in the dataset?

```{r}
table(subject$infancy_vac)
```

> Q5. How many Male and Female subjects/patients are in the dataset?

```{r}
table(subject$biological_sex)
```

> Q6. What is the breakdown of race and biological sex (e.g. number of Asian females, White males etc…)?

```{r}
table(subject$race, subject$biological_sex)
```

This is not representative of the US population, but it's the best we have.

Let's obtain more data from CMI-PB:

```{r}
specimen <- read_json("http://cmi-pb.org/api/v5_1/specimen", 
                      simplifyVector=TRUE)

ab_data <- read_json("http://cmi-pb.org/api/v5_1/plasma_ab_titer", 
                     simplifyVector=TRUE)
```

```{r}
head(specimen)
```

```{r}
head(ab_data)
```

I now have 3 tables of data from CMI-PB: `subject`, `specimen`, and `ab_data`. I need to join these tables so I will have all the info I need to work with.  

For this we will use the `inner_join` function from the **dplyr** package.

```{r}
library(dplyr)

meta <- inner_join(subject, specimen)
head(meta)
```

Now we can join our `ab_data` table to `meta`so we have all 3 sets of information in the same place. 

```{r}
abdata<- inner_join(meta, ab_data)
head(abdata)
dim(abdata)
```

Q. How many different antibody isotypes are there is this dataset?

```{r}
length(abdata$isotype)
```

```{r}
table(abdata$isotype)
```

```{r}
table(abdata$antigen)
```

I want a plot of antigen levels across the whole dataset. 

```{r}
ggplot(abdata, aes(MFI, antigen))+
  geom_boxplot()
```

```{r}
ggplot(abdata, aes(MFI_normalised, antigen))+
  geom_boxplot()
```

Antigens like FIM2/3, PT, FELD1 have quite a large range of values. Others, like Measles don't show much activity. 

> Q. Are there differences at this whole-data level between aP and wP?

```{r}
ggplot(abdata, aes(MFI_normalised, antigen, col=infancy_vac))+
  geom_boxplot()
```

This is a lot and we're looking at every antibody type. 

```{r}
ggplot(abdata, aes(MFI_normalised, antigen))+
  geom_boxplot()+
  facet_wrap(~infancy_vac)
```

## Examine just IgG levels

Isolate just isotype IgG.

```{r}
igg <- abdata %>% filter(isotype == "IgG")
head(igg)
```

```{r}
ggplot(igg, aes(MFI_normalised, antigen, col=infancy_vac))+
  geom_boxplot()
```

This is still too muddy, across all time points before and after vaccination. Let's dig into a more specific point, time course of IgG isotype PT antigen levels across aP and wP individuals. First we filtered to include the 2021 dataset, then filtered to look at IgG and PT data only. We plotted this data and colored by infancy_vac, with a setup of wP vs aP. 

```{r}
abdata.21 <- abdata %>% filter(dataset == "2021_dataset")

abdata.21 %>% 
  filter(isotype == "IgG",  antigen == "PT") %>%
  
  ggplot() +
    aes(x=planned_day_relative_to_boost,
        y=MFI_normalised,
        col=infancy_vac,
        group=subject_id) +
    geom_point() +
    geom_line() +
    geom_vline(xintercept=0, linetype="dashed") +
    geom_vline(xintercept=14, linetype="dashed") +
  labs(title="2021 dataset IgG PT",
       subtitle = "Dashed lines indicate day 0 (pre-boost) and 14 (apparent peak levels)")
```

