---
title: "Comparative structure analysis"
author: "Dylan Mullaney (A16869792)"
format: html
---

Here we run through a complete "pipeline" of structure analysisthat begins with a single sequence identifier and ends in a PCA analysis. 

```{r}
library(bio3d)
library(httr)
```

## Step 1. Get a seqeunce
Retrieve a sequence for the protein we are interested in. We will take ADK "1ake_A".

```{r}
id <-"1ake_A"
aa <- get.seq(id)
aa
```

## Step 2. BLAST 

Run a BLAST search of the PDB for all related sequecnes to our input 'aa'

```{r}
blast <- blast.pdb(aa)
```

```{r}
hits <- plot(blast)
```

## Step 3. Download all structures

These are our 'top hits' i.e. all the structure in teh PDB database realted to our input sequence.

```{r}
hits$pdb.id
```

Dowload related PDB files, use this to save to computer files and then view on molstar

```{r}
files <- get.pdb(hits$pdb.id, path="pdbs", split=TRUE, gzip=TRUE)
```



## Step 4. Align related sequences and superpose

```{r}
pdbs <- pdbaln(files, fit = TRUE, exefile="msa")
```

```{r}
#pdbs
```

## Step. 5 PCA 

Let's use our old friend PCA to make sense of these confusing, complicated structure relationships. 
```{r}
pc <- pca(pdbs)
plot(pc)
```

It appears that PC1 accounts for more than 90% of the data. Let's make a trajectory (or movie) of the main conformational changes captured by PC1. We will use the `mktrj()` function for this.

```{r}
mktrj(pc, file= "pca_results.pdb")
```

What if you can't find a structure anywhere for something you want to analyze? You can generate your own structure predictions with ColabFold, looking at the monomers and dimers.










